---
title: "final"
format: html
---
repo: https://github.com/wywydubs/ENVS-193DS_spring-2025_final
```{r set up, message=FALSE, warning=FALSE}
library(tidyverse) # Load packages (based on hw 3 packages)
library(here)
library(gt) # chose gt

library(janitor) # Optional packages
library(readxl)
library(readr)
library(DHARMa)
library(MuMIn)
sst <- read_csv(here("data", "SST_update2023.csv")) # read in data for prob 2 
nest_boxes <- read_csv(here("data", "occdist.csv")) |>
  clean_names() # read in data for prob 3 
```



# Problem 1

## a. Transparent statistical methods

In part 1 they used some type of correlation test. In part two they used an 
ANOVA test. 

## b. More information needed

Effect size could be helpful in this context, as it would tell us more about
the strength of the data, which would give more real world application (ex.
a large effect size would explain a substantial nitrogen load difference. 
Actually presenting the means and their standard deviations in some way would 
also help, as this would give a digestible way to actually view the data, not 
just the conclusion from the data (ex. mean fertilizer samples had x amount of 
nitrogen present). 


## c. Suggestions for Rewriting

1. We found significant correlation between distance from headwater (km) 
and annual total nitrogen load (kg year^-1), telling us that there is a
relationship between the two likely due to impacts from the runoff surrounding 
the headwaters. correlation test (pearson?) (r = correlation 
coefficient, p = 0.03, α = significance level)

2. The source of the runoff has a statistically significant effect on the
average annual nitrogen load, telling us that nitrogen loads vary based on
where runoff is coming from (urban land, atmospheric deposition, fertilizer, 
wastewater treatment, and grasslands). (ANOVA: F = F-statistic, p = 0.02, η² = 
effect size, α = significance level).


# Problem 2

## a. data cleaning and summarizing 

```{r cleaning, message=FALSE, warning=FALSE}
sst_clean <- sst |> # rename
  mutate(
    year  = year(date),
    month = month(date)) |>  # create new columns
      
    filter(year >= 2018)|>  # filter to match structure
  group_by(year, month) |> # group by new columns 
  summarize(
    mean_monthly_sst = mean(temp), # add in mean 
    .groups = 'drop') |> # make output ungrouped 
  # (ChatGPT troubleshooting suggestion)
  
  mutate(
    year = factor(year), # convert column to ordered month list (abbreviated)
     month = factor(month, levels = 1:12, labels = month.abb, ordered = TRUE) 
  )
  
 slice_sample(sst_clean, n = 5) # give the 5 samples

str(sst_clean) # shwo the structure

```




# b. Data visualization

```{r recreation, message=FALSE, warning=FALSE}
ggplot(
  data = sst_clean, # specify in data
  aes(x = month, y = mean_monthly_sst, group = year, color = year)) + 
  # define x, y, groups, color
  geom_line(linewidth = 0.7) + # add in line and points
  geom_point(size = 2) + 
  labs(
    x = "Month",
    y = "Mean monthly sea surface temperature (°C)",
    color = "Year") + # add in names for x and y and color
  scale_color_brewer(palette = "Blues") + # add in blues colors
  theme_bw() + # assign correct theme
  theme(
    panel.grid.major = element_blank(), # get rid of grid lines
    panel.grid.minor = element_blank(),
    legend.position = c(0.15, 0.75) # rough legend position
  )
```



  # Problem 3 
  
 ## a. Response Variable
  
  In this data set, 1s indicate that a nest box was occupied by that specific 
  bird species, and 0s indicate that the nest box was unoccupied. Biologically, 
  this represents if a certain nest box provided viable nesting habitat for 
  a specific species during the breeding season.
  
  
 ## b. Purpose of Study
  
  The Swift Parrots are the main bird intended to be protected, while the 
  Starlings and Martins are the competitors. Since the Swift Parrots are in a
  more vulnerable population outlook, they require more protection, while the 
  other two species are not endangered currently. 
  
 ## c. Difference in “seasons”
  
  The two seasons are the breeding years 2016 and 2019. In this study, 
  2016 was for new nest boxes, and 2019 was for established boxes 
  that had been there for 3 years which gave researchers the ability to compare
  initial inhabitant data to established box inhabitant data. 
  
 ## d. Table 
  
```{r table of models}

model_table <- tibble::tibble( # make data frame
  `Model #` = 1:4, # assign numbers
  `Season Variable Present?` = c("No", "Yes", "No", "Yes"), # add season column
  `Distance to Forest Edge Present?` = c("No", "No", "Yes", "Yes"), 
  # add forest column
  
  `Model Description` = c(
    "Null model: assumes no effect from the variables.",
    "Season-only model: tests if year/season affects occupancy.",
    "Distance-only model: tests if distance to forest affects occupancy.",
    "Saturated model: includes interaction between season and distance 
    variables."
  ) # add in descriptions
)

model_table |> #create a table
  gt() |> 
  tab_header(
    title = "Table of Models for Nest Box Occupancy" # add title in header
  )



```
  
  
## e. Run Models

```{r models, message=FALSE, warning=FALSE}

mod1 <- glm(e ~ 1, data = nest_boxes, family = binomial) # create null model

mod2 <- glm(e ~ season, data = nest_boxes, family = binomial) 
# create season model

mod3 <- glm(e ~ edge_distance, data = nest_boxes, family = binomial) 
# create edge distance model 

mod4 <- glm(e ~ season * edge_distance, data = nest_boxes, family = binomial)
# create saturated model
```

## f. Check the diagnostics 
```{r simulate, message=FALSE, warning=FALSE}

# simulate and plot for all models 

res1 <- simulateResiduals(mod1)
plot(res1)

res2 <- simulateResiduals(mod2)
plot(res2)

res3 <- simulateResiduals(mod3)
plot(res3)

res4 <- simulateResiduals(mod4)
plot(res4)

```


## g. Select the best model
```{r MuMin}
options(na.action = "na.fail") # troubleshooting line from MuMin guide

# Use dredge to give every possible model from our model 4 with all variables
dredge_results <- dredge(mod4)

# Show the top models ranked by AICc (supposedly more accurate vs AIC)
print(dredge_results)
```

The best model as determined by Akaike’s Information Criterion corrected for 
small sample size (AICc) was the model that measured the occupancy based on
the relationship between edge distance and season. 
